---
title: "Raster Processing: terra & GDAL"
author: "Chantal Fuchs"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
engine: knitr
---

::: {.callout-note}
**Declaration of AI Use:** Parts of this submission were developed with the
assistance of Claude (Anthropic). The AI was used to help with code structure,
debugging, and R syntax. All results were verified by me.
:::

---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(terra)
library(dplyr)
library(tidyr)
library(tictoc)
library(knitr)
library(ggplot2)
library(purrr)

tictoc::tic.clearlog()

path_tlm   <- "data/SWISSTLM3D_2026_LV95_LN02.gpkg"
path_bound <- "data/swissBOUNDARIES3D_1_5_LV95_LN02.gpkg"

layer_cantons <- "tlm_kantonsgebiet"
layer_forest  <- "tlm_bb_bodenbedeckung"
```

---

## Task 1 — Forest per Canton with `terra`

This is the same analysis as in Vector Deepdive — calculating forest cover per
canton — but using the `terra` package instead of `sf`. The main difference is
that `terra` works with `SpatVector` objects instead of `sf` objects, and it
calls the underlying C++ code more directly, which is supposed to be faster.

### Step 1: Load canton boundaries

```{r}
#| label: terra-step1

tictoc::tic(msg = "terra | Step 1: Load canton boundaries")

cantons <- terra::vect(path_bound, layer = layer_cantons)

tictoc::toc(log = TRUE)

cat("CRS:              ", terra::crs(cantons, describe = TRUE)$code, "\n")
cat("Number of cantons:", nrow(cantons), "\n")
```

### Step 2: Load and filter forest polygons

```{r}
#| label: terra-step2

tictoc::tic(msg = "terra | Step 2: Load and filter forest polygons")

bb     <- terra::vect(path_tlm, layer = layer_forest)
forest <- bb[bb$objektart == "Wald", ]

tictoc::toc(log = TRUE)

cat("Forest polygon count:", nrow(forest), "\n")
```

### Step 3: Check and align CRS

```{r}
#| label: terra-step3

tictoc::tic(msg = "terra | Step 3: Verify CRS alignment")

crs_cantons <- terra::crs(cantons, describe = TRUE)$code
crs_forest  <- terra::crs(forest,  describe = TRUE)$code

cat("Canton CRS:", crs_cantons, "\n")
cat("Forest CRS:", crs_forest,  "\n")
cat("CRS match: ",  crs_cantons == crs_forest, "\n")

if (crs_cantons != crs_forest) {
  forest <- terra::project(forest, terra::crs(cantons))
}

tictoc::toc(log = TRUE)
```

### Step 4: Spatial intersection

```{r}
#| label: terra-step4

tictoc::tic(msg = "terra | Step 4: Intersect forest with cantons")

forest_canton <- terra::intersect(forest, cantons)

tictoc::toc(log = TRUE)

cat("Intersection result rows:", nrow(forest_canton), "\n")
```

### Step 5: Calculate forest percentage per canton

```{r}
#| label: terra-step5

tictoc::tic(msg = "terra | Step 5: Calculate forest % per canton")

forest_canton$forest_area_m2 <- terra::expanse(forest_canton, unit = "m")

forest_sum <- as.data.frame(forest_canton) |>
  group_by(name) |>
  summarise(forest_area_m2 = sum(forest_area_m2, na.rm = TRUE), .groups = "drop")

cantons_df <- as.data.frame(cantons)
cantons_df$canton_area_m2 <- terra::expanse(cantons, unit = "m")

result_terra <- cantons_df |>
  select(name, canton_area_m2) |>
  left_join(forest_sum, by = "name") |>
  mutate(
    forest_area_m2 = replace_na(forest_area_m2, 0),
    forest_pct     = (forest_area_m2 / canton_area_m2) * 100
  ) |>
  arrange(desc(forest_pct))

tictoc::toc(log = TRUE)
```

### Results

```{r}
#| label: terra-table

kable(
  result_terra |>
    mutate(
      `Canton Area (km²)` = round(canton_area_m2 / 1e6, 1),
      `Forest Area (km²)` = round(forest_area_m2 / 1e6, 1),
      `Forest (%)`        = round(forest_pct, 2)
    ) |>
    select(Canton = name,
           `Canton Area (km²)`,
           `Forest Area (km²)`,
           `Forest (%)`),
  caption = "Forest coverage per Swiss canton (terra)"
)
```

```{r}
#| label: terra-plot

ggplot(result_terra,
       aes(x = reorder(name, forest_pct), y = forest_pct)) +
  geom_col(fill = "#2d6a4f") +
  coord_flip() +
  labs(
    title    = "Forest Coverage per Swiss Canton",
    subtitle = "Calculated with terra",
    x        = NULL,
    y        = "Forest (%)"
  ) +
  theme_minimal(base_size = 11)
```

---

## Task 2 — Forest per Canton with GDAL

Same analysis again, but using GDAL this time. On Windows, the GDAL command-line
tools (`ogr2ogr`) are not available directly in the terminal, so I used
`sf::gdal_utils()` instead — this calls the same underlying GDAL C library.
For the timing, I wrap each GDAL step with `tictoc::tic()` and `tictoc::toc()`
in R chunks around the GDAL calls, as suggested in the assignment.

### Step 1: Export canton boundaries via GDAL

```{r}
#| label: gdal-step1

tictoc::tic(msg = "gdal | Step 1: Export cantons via GDAL (ogr2ogr)")

sf::gdal_utils(
  util        = "vectortranslate",
  source      = path_bound,
  destination = "data/cantons_gdal.gpkg",
  options     = c(
    "-f", "GPKG",
    "-nln", "kantonsgebiet",
    "-sql", paste0("SELECT * FROM ", layer_cantons),
    "-overwrite"
  )
)

tictoc::toc(log = TRUE)
cat("cantons_gdal.gpkg created.\n")
```

### Step 2: Extract forest polygons via GDAL

```{r}
#| label: gdal-step2

tictoc::tic(msg = "gdal | Step 2: Extract forest polygons via GDAL")

sf::gdal_utils(
  util        = "vectortranslate",
  source      = path_tlm,
  destination = "data/forest_gdal.gpkg",
  options     = c(
    "-f", "GPKG",
    "-sql", paste0(
      "SELECT * FROM ", layer_forest,
      " WHERE objektart = 'Wald'"
    ),
    "-overwrite"
  )
)

tictoc::toc(log = TRUE)
cat("forest_gdal.gpkg created.\n")
```

### Step 3: Spatial intersection

```{r}
#| label: gdal-step3

tictoc::tic(msg = "gdal | Step 3: Spatial intersection via sf (GDAL backend)")

forest_gdal  <- sf::st_read("data/forest_gdal.gpkg",  quiet = TRUE)
cantons_gdal <- sf::st_read("data/cantons_gdal.gpkg", quiet = TRUE)

forest_canton_gdal <- sf::st_intersection(forest_gdal, cantons_gdal)
forest_canton_gdal$forest_area_m2 <- as.numeric(sf::st_area(forest_canton_gdal))

sf::st_write(forest_canton_gdal, "data/forest_canton_gdal.gpkg",
             delete_dsn = TRUE, quiet = TRUE)

tictoc::toc(log = TRUE)
cat("Intersection rows:", nrow(forest_canton_gdal), "\n")
```

### Step 4: Aggregate per canton

```{r}
#| label: gdal-step4

tictoc::tic(msg = "gdal | Step 4: Aggregate forest area per canton")

forest_by_canton_gdal <- forest_canton_gdal |>
  sf::st_drop_geometry() |>
  group_by(name) |>
  summarise(forest_area_m2 = sum(forest_area_m2, na.rm = TRUE), .groups = "drop")

tictoc::toc(log = TRUE)
cat("Cantons aggregated:", nrow(forest_by_canton_gdal), "\n")
```

### Step 5: Calculate percentages in R

```{r}
#| label: gdal-step5

tictoc::tic(msg = "gdal | Step 5: Join and compute percentages in R")

cantons_sf <- sf::st_read("data/cantons_gdal.gpkg", quiet = TRUE)
cantons_sf$canton_area_m2 <- as.numeric(sf::st_area(cantons_sf))

result_gdal <- cantons_sf |>
  sf::st_drop_geometry() |>
  select(name, canton_area_m2) |>
  left_join(forest_by_canton_gdal, by = "name") |>
  mutate(
    forest_area_m2 = replace_na(forest_area_m2, 0),
    forest_pct     = (forest_area_m2 / canton_area_m2) * 100
  ) |>
  arrange(desc(forest_pct))

tictoc::toc(log = TRUE)
```

```{r}
#| label: gdal-table

kable(
  result_gdal |>
    mutate(
      `Canton Area (km²)` = round(canton_area_m2 / 1e6, 1),
      `Forest Area (km²)` = round(forest_area_m2 / 1e6, 1),
      `Forest (%)`        = round(forest_pct, 2)
    ) |>
    select(Canton = name,
           `Canton Area (km²)`,
           `Forest Area (km²)`,
           `Forest (%)`),
  caption = "Forest coverage per Swiss canton (GDAL)"
)
```

---

## Task 3 — Execution Time Comparison

```{r}
#| label: timing-parse

log_raw   <- tictoc::tic.log(format = FALSE)

timing_df <- map_df(log_raw, function(x) {
  tibble(
    step     = x$msg,
    time_sec = as.numeric(x$toc - x$tic)
  )
}) |>
  mutate(
    approach = case_when(
      startsWith(step, "terra") ~ "terra",
      startsWith(step, "gdal")  ~ "GDAL",
      TRUE                      ~ "other"
    )
  )

kable(
  timing_df |>
    select(Step = step, `Time (s)` = time_sec, Approach = approach) |>
    mutate(`Time (s)` = round(`Time (s)`, 3)),
  caption = "Execution time per step"
)
```

```{r}
#| label: timing-totals

terra_total <- sum(timing_df$time_sec[timing_df$approach == "terra"])
gdal_total  <- sum(timing_df$time_sec[timing_df$approach == "GDAL"])

# Times from previous weeks (sf and DuckDB from Vector Deepdive)
sf_total     <- 237.94
duckdb_total <- 191.14

comparison_df <- tibble(
  Approach  = c("sf", "DuckDB", "terra", "GDAL"),
  total_sec = c(sf_total, duckdb_total, terra_total, gdal_total)
)

kable(
  comparison_df |>
    arrange(total_sec) |>
    mutate(total_sec = round(total_sec, 2)),
  col.names = c("Approach", "Total time (s)"),
  caption   = "Total execution time per approach"
)
```

```{r}
#| label: timing-plot

ggplot(
  comparison_df,
  aes(x = reorder(Approach, total_sec), y = total_sec, fill = Approach)
) +
  geom_col(show.legend = FALSE, width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c(
    "sf"     = "#4a90d9",
    "DuckDB" = "#f5a623",
    "terra"  = "#2d6a4f",
    "GDAL"   = "#c0392b"
  )) +
  labs(
    title    = "Total Execution Time by Approach",
    subtitle = "Same task (forest % per canton) — four different tools",
    x        = NULL,
    y        = "Total time (seconds)"
  ) +
  theme_minimal(base_size = 12)
```

### Discussion

Going into this task I expected GDAL to be the fastest, since it is a low-level
C library that runs directly without much overhead. The results showed something
different though: `terra` was actually the fastest overall, while GDAL ended up
being the slowest of the four approaches.

After thinking about it, this makes sense for how the GDAL approach was set up
here. On Windows, the command-line `ogr2ogr` tool is not available, so I used
`sf::gdal_utils()` instead. The intersection step still went through
`sf::st_intersection()`, which means it was essentially running `sf` code anyway.
So the GDAL pipeline in this case was not purely GDAL — it was a mix, which
explains why it did not perform as expected.

`terra` was the most positive surprise. The syntax is similar to `sf` but it
handles the data differently internally — it uses `SpatVector` objects and
processes things in C++ more directly. Getting used to `SpatVector` instead of
`sf` objects was a bit confusing at first, but once the code was running it was
noticeably quicker.

One thing I noticed is that the results from all four approaches are very
similar numerically, which is reassuring — even though the tools work differently,
they all end up with the same answer. The choice of tool matters mainly for
performance, not correctness.

If I were doing this again with a much larger dataset, I would probably use
`terra` or a proper GDAL command-line setup (on Linux or Mac), since those
would likely show the biggest advantage at scale.

**Note on GDAL setup:** The assignment suggests using `{bash}` chunks with `ogr2ogr` directly. On Windows, GDAL command-line tools are not available by default, so I used `sf::gdal_utils()` as an equivalent alternative, which calls the same underlying GDAL C library.