---
title: "Raster Processing: terra & GDAL"
author: "Chantal"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
engine: knitr
---

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

library(sf)
library(terra)
library(dplyr)
library(tidyr)
library(tictoc)
library(knitr)
library(ggplot2)
library(purrr)

# Clear tictoc log at the start
tictoc::tic.clearlog()

# Data paths
path_tlm   <- "data/SWISSTLM3D_2026_LV95_LN02.gpkg"
path_bound <- "data/swissBOUNDARIES3D_1_5_LV95_LN02.gpkg"

# Layer names
layer_cantons <- "tlm_kantonsgebiet"
layer_forest  <- "tlm_bb_bodenbedeckung"
```

---

## Task 1 — Forest per Canton with `terra`

We repeat the forest-per-canton analysis from Vector Deepdive, this time using the
`terra` package instead of `sf`. `terra` calls GDAL/GEOS internally via compiled
C++ code, which reduces R overhead and is typically faster than `sf` for large datasets.

### Step 1: Load canton boundaries

```{r}
#| label: terra-step1

tictoc::tic(msg = "terra | Step 1: Load canton boundaries")

cantons <- terra::vect(path_bound, layer = layer_cantons)

tictoc::toc(log = TRUE)

cat("CRS:              ", terra::crs(cantons, describe = TRUE)$code, "\n")
cat("Number of cantons:", nrow(cantons), "\n")
```

### Step 2: Load and filter forest polygons

```{r}
#| label: terra-step2

tictoc::tic(msg = "terra | Step 2: Load and filter forest polygons")

bb     <- terra::vect(path_tlm, layer = layer_forest)
forest <- bb[bb$objektart == "Wald", ]

tictoc::toc(log = TRUE)

cat("Forest polygon count:", nrow(forest), "\n")
```

### Step 3: Check and align CRS

Both datasets are already in LV95 (CH1903+ / EPSG:2056), so no reprojection is needed.
We verify this step nonetheless and measure the time.

```{r}
#| label: terra-step3

tictoc::tic(msg = "terra | Step 3: Verify CRS alignment")

crs_cantons <- terra::crs(cantons, describe = TRUE)$code
crs_forest  <- terra::crs(forest,  describe = TRUE)$code

cat("Canton CRS:", crs_cantons, "\n")
cat("Forest CRS:", crs_forest,  "\n")
cat("CRS match: ",  crs_cantons == crs_forest, "\n")

# Reproject only if needed (not the case here)
if (crs_cantons != crs_forest) {
  forest <- terra::project(forest, terra::crs(cantons))
}

tictoc::toc(log = TRUE)
```

### Step 4: Spatial intersection

```{r}
#| label: terra-step4

tictoc::tic(msg = "terra | Step 4: Intersect forest with cantons")

# terra::intersect() clips forest polygons to canton boundaries
# and appends canton attributes to each resulting polygon
forest_canton <- terra::intersect(forest, cantons)

tictoc::toc(log = TRUE)

cat("Intersection result rows:", nrow(forest_canton), "\n")
```

### Step 5: Calculate forest percentage per canton

```{r}
#| label: terra-step5

tictoc::tic(msg = "terra | Step 5: Calculate forest % per canton")

# Area of each clipped forest polygon (m²)
forest_canton$forest_area_m2 <- terra::expanse(forest_canton, unit = "m")

# Sum forest area per canton (group by canton name)
forest_sum <- as.data.frame(forest_canton) |>
  group_by(name) |>
  summarise(forest_area_m2 = sum(forest_area_m2, na.rm = TRUE), .groups = "drop")

# Canton total areas
cantons_df <- as.data.frame(cantons)
cantons_df$canton_area_m2 <- terra::expanse(cantons, unit = "m")

# Join and compute percentage
result_terra <- cantons_df |>
  select(name, canton_area_m2) |>
  left_join(forest_sum, by = "name") |>
  mutate(
    forest_area_m2 = replace_na(forest_area_m2, 0),
    forest_pct     = (forest_area_m2 / canton_area_m2) * 100
  ) |>
  arrange(desc(forest_pct))

tictoc::toc(log = TRUE)
```

### Results

```{r}
#| label: terra-table

kable(
  result_terra |>
    mutate(
      `Canton Area (km2)` = round(canton_area_m2 / 1e6, 1),
      `Forest Area (km2)` = round(forest_area_m2 / 1e6, 1),
      `Forest (%)`        = round(forest_pct, 2)
    ) |>
    select(Canton = name,
           `Canton Area (km2)`,
           `Forest Area (km2)`,
           `Forest (%)`),
  caption = "Forest coverage per Swiss canton (terra)"
)
```

```{r}
#| label: terra-plot

ggplot(result_terra,
       aes(x = reorder(name, forest_pct), y = forest_pct)) +
  geom_col(fill = "#2d6a4f") +
  coord_flip() +
  labs(
    title    = "Forest Coverage per Swiss Canton",
    subtitle = "Calculated with terra",
    x        = NULL,
    y        = "Forest (%)"
  ) +
  theme_minimal(base_size = 11)
```

---

## Task 2 — Forest per Canton with GDAL

We repeat the same analysis using GDAL. On Windows, the GDAL command-line tools
are not available in the system terminal by default. We therefore use `sf::gdal_utils()`,
which is the official R interface to the GDAL library bundled with the `sf` package —
this calls the exact same underlying GDAL C library as `ogr2ogr` on the command line.

The timing workaround (tictoc in R chunks around GDAL calls) is the same approach
as shown in the assignment for bash chunks.

### Step 1: Reproject / export canton boundaries

```{r}
#| label: gdal-step1

tictoc::tic(msg = "gdal | Step 1: Export cantons via GDAL (ogr2ogr)")

# gdal_utils("vectortranslate") is equivalent to ogr2ogr on the command line
sf::gdal_utils(
  util      = "vectortranslate",
  source    = path_bound,
  destination = "data/cantons_gdal.gpkg",
  options   = c(
    "-f", "GPKG",
    "-nln", "kantonsgebiet",
    "-sql", paste0("SELECT * FROM ", layer_cantons),
    "-overwrite"
  )
)

tictoc::toc(log = TRUE)
cat("cantons_gdal.gpkg created.\n")
```

### Step 2: Extract forest polygons

```{r}
#| label: gdal-step2

tictoc::tic(msg = "gdal | Step 2: Extract forest polygons via GDAL")

sf::gdal_utils(
  util        = "vectortranslate",
  source      = path_tlm,
  destination = "data/forest_gdal.gpkg",
  options     = c(
    "-f", "GPKG",
    "-sql", paste0(
      "SELECT * FROM ", layer_forest,
      " WHERE objektart = 'Wald'"
    ),
    "-overwrite"
  )
)

tictoc::toc(log = TRUE)
cat("forest_gdal.gpkg created.\n")
```

### Step 3: Spatial intersection

```{r}
#| label: gdal-step3

tictoc::tic(msg = "gdal | Step 3: Spatial intersection via sf (GDAL backend)")

# Load both layers
forest_gdal  <- sf::st_read("data/forest_gdal.gpkg",  quiet = TRUE)
cantons_gdal <- sf::st_read("data/cantons_gdal.gpkg", quiet = TRUE)

# Intersection (sf uses GDAL/GEOS internally)
forest_canton_gdal <- sf::st_intersection(forest_gdal, cantons_gdal)
forest_canton_gdal$forest_area_m2 <- as.numeric(sf::st_area(forest_canton_gdal))

# Save for next step
sf::st_write(forest_canton_gdal, "data/forest_canton_gdal.gpkg", 
             delete_dsn = TRUE, quiet = TRUE)

tictoc::toc(log = TRUE)
cat("Intersection rows:", nrow(forest_canton_gdal), "\n")
```

### Step 4: Aggregate per canton

```{r}
#| label: gdal-step4

tictoc::tic(msg = "gdal | Step 4: Aggregate forest area per canton")

# Aggregate directly in R using the result from step 3
forest_by_canton_gdal <- forest_canton_gdal |>
  sf::st_drop_geometry() |>
  group_by(name) |>
  summarise(forest_area_m2 = sum(forest_area_m2, na.rm = TRUE), .groups = "drop")

tictoc::toc(log = TRUE)
cat("Cantons aggregated:", nrow(forest_by_canton_gdal), "\n")
```

### Step 5: Calculate percentages in R

```{r}
#| label: gdal-step5

tictoc::tic(msg = "gdal | Step 5: Join and compute percentages in R")

cantons_sf <- sf::st_read("data/cantons_gdal.gpkg", quiet = TRUE)
cantons_sf$canton_area_m2 <- as.numeric(sf::st_area(cantons_sf))

result_gdal <- cantons_sf |>
  sf::st_drop_geometry() |>
  select(name, canton_area_m2) |>
  left_join(forest_by_canton_gdal, by = "name") |>
  mutate(
    forest_area_m2 = replace_na(forest_area_m2, 0),
    forest_pct     = (forest_area_m2 / canton_area_m2) * 100
  ) |>
  arrange(desc(forest_pct))

tictoc::toc(log = TRUE)
```

```{r}
#| label: gdal-table

kable(
  result_gdal |>
    mutate(
      `Canton Area (km2)` = round(canton_area_m2 / 1e6, 1),
      `Forest Area (km2)` = round(forest_area_m2 / 1e6, 1),
      `Forest (%)`        = round(forest_pct, 2)
    ) |>
    select(Canton = name,
           `Canton Area (km2)`,
           `Forest Area (km2)`,
           `Forest (%)`),
  caption = "Forest coverage per Swiss canton (GDAL)"
)
```

---

## Task 3 — Execution Time Comparison

```{r}
#| label: timing-parse

log_raw   <- tictoc::tic.log(format = FALSE)

timing_df <- map_df(log_raw, function(x) {
  tibble(
    step     = x$msg,
    time_sec = as.numeric(x$toc - x$tic)
  )
}) |>
  mutate(
    approach = case_when(
      startsWith(step, "terra") ~ "terra",
      startsWith(step, "gdal")  ~ "GDAL",
      TRUE                      ~ "other"
    )
  )

kable(
  timing_df |>
    select(Step = step, `Time (s)` = time_sec, Approach = approach) |>
    mutate(`Time (s)` = round(`Time (s)`, 3)),
  caption = "Execution time per step"
)
```

```{r}
#| label: timing-totals

terra_total <- sum(timing_df$time_sec[timing_df$approach == "terra"])
gdal_total  <- sum(timing_df$time_sec[timing_df$approach == "GDAL"])

# ── Insert your measured times from previous weeks here (in seconds) ──────────
sf_total     <- 325.12   #  from your sf assignment

duckdb_total <- 243.29   #  from your DuckDB assignment
# ─────────────────────────────────────────────────────────────────────────────

comparison_df <- tibble(
  Approach  = c("GeoPandas", "DuckDB", "terra", "GDAL"),
  total_sec = c(sf_total, duckdb_total, terra_total, gdal_total)
)

kable(
  comparison_df |>
    filter(!is.na(total_sec)) |>
    arrange(total_sec) |>
    mutate(total_sec = round(total_sec, 2)),
  col.names = c("Approach", "Total time (s)"),
  caption   = "Total execution time per approach"
)
```

```{r}
#| label: timing-plot

ggplot(
  comparison_df |> filter(!is.na(total_sec)),
  aes(x = reorder(Approach, total_sec), y = total_sec, fill = Approach)
) +
  geom_col(show.legend = FALSE, width = 0.6) +
  coord_flip() +
  scale_fill_manual(values = c(
    "GeoPandas"     = "#4a90d9",
    "DuckDB" = "#f5a623",
    "terra"  = "#2d6a4f",
    "GDAL"   = "#c0392b"
  )) +
  labs(
    title    = "Total Execution Time by Approach",
    subtitle = "Same task (forest % per canton) — four different tools",
    x        = NULL,
    y        = "Total time (seconds)"
  ) +
  theme_minimal(base_size = 12)
```

### Discussion

**`sf`** loads data fully into R memory and processes it with GEOS. It is the most
convenient for R users (tidy-compatible), but typically the slowest for large datasets
because all computation happens in R session memory.

**`DuckDB`** pushes spatial operations into a compiled SQL query engine that runs
in parallel. It avoids loading the full dataset into R memory upfront, which makes
it much faster than `sf` for large inputs.

**`terra`** is designed for raster data but also handles vectors via `SpatVector`.
It calls GDAL/GEOS through compiled C++ internally, which removes most R overhead
and makes it noticeably faster than `sf`.

**`GDAL`** (here via `sf::gdal_utils()`, which calls the same underlying C library
as `ogr2ogr` on the command line) works directly at the library level with minimal
overhead. It can stream data without loading everything into RAM, making it very
efficient for large files. The trade-off is more verbose syntax.

**Expected ranking (slowest to fastest):**  
`sf` → `terra` → `GDAL` ≈ `DuckDB`

The exact results depend on dataset size and hardware, but compiled/query-engine
approaches consistently outperform pure-R solutions as data volume grows.
